### 2022-10-26 draft/draft.tex
minor tweak to proof of theorem 4.2
minor updates
adding in detaisls on bezier curve fitting

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
To illustrate this, consider the typical approach to building a statistical or machine learning model.
Data are often represented as an \(n \times p\) matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^p\) and then assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
Instead of feature vector \(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p\), we observe \(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Let the size of the graph, \(n\), be fixed.
  Then each vertex \(v_i\), and its corresponding label, \(z_i\), are also fixed, in addition to the pairwise community edge probabilities \(\{\theta_{k \ell}\}_K\), which are all treated as model parameters.
  Under this sampling scheme, there is no notion of increasing \(n\).
\item
  The vertices and edges are sampled in sequence, and we suppose that each vertex \(v_i\) comes with a corresponding label, \(z_i\), which determines its edge probabilities to the other vertices.
  Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with \(\{\theta_{k \ell}\}_K\), but there is a notion of increasing sample size \(n\), which allows for asymptotics.
\item
  The vertices are sampled in sequence, and each label, \(z_i\), is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
  The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top\) and \(\{\theta_{k \ell}\}_K\).
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(t_1, ..., t_n \iid F\) for probability distribution \(F\) with support \([0, 1]\).
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), the community labels.
\item
  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\item
  Draw \(A \sim \GRDPG_{p,q}(X; \rho_n)\).
\end{enumerate}

\begin{example}
\label{ex:intersect-curves}
Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$. 
This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
The community structure is also not clear from the adjacency matrix. 
However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).

@@

For a simulation study, we again turn to examples. \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
The exact simulation setup is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(1/2, 1/2)\).
\item
  Draw \(t_1, ..., t_n \iid \Uniform(0, 1)\).
\item
  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top\) and \(g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top\).
  Collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\item
  Draw \(A \sim \RDPG(X; \rho_n)\) (for these simulations we fix \(\rho_n \equiv 1\)).
\end{enumerate}

For this simulation study, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community.
For each setting, we simulate 50 replicates.
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(0)} = p_2^{(0)} = (0, 0)\).
Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting.
The simulation results suggest that for large \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.

@@

\caption{Median and IQR error rates over 50 replicates.}\label{fig:sim-curves}
\end{figure}

@@

@@

>>>>

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
To illustrate this, consider the typical approach to building a statistical or machine learning model.
Data are often represented as an \(n \times p\) matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^p\) and then assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
Instead of feature vector \(x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p\), we observe \(a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Let the size of the graph, \(n\), be fixed.
  Then each vertex \(v_i\), and its corresponding label, \(z_i\), are also fixed, in addition to the pairwise community edge probabilities \(\{\theta_{k \ell}\}_K\), which are all treated as model parameters.
  Under this sampling scheme, there is no notion of increasing \(n\).
\item
  The vertices and edges are sampled in sequence, and we suppose that each vertex \(v_i\) comes with a corresponding label, \(z_i\), which determines its edge probabilities to the other vertices.
  Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with \(\{\theta_{k \ell}\}_K\), but there is a notion of increasing sample size \(n\), which allows for asymptotics.
\item
  The vertices are sampled in sequence, and each label, \(z_i\), is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\).
  The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}_K\).
  Under this sampling scheme, both the edges and the labels are random, with fixed parameters \(\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top\) and \(\{\theta_{k \ell}\}_K\).
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(t_1, ..., t_n \iid F\) for probability distribution \(F\) with support \([0, 1]\).
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), the community labels.
\item
  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
\item
  Draw \(A \sim \GRDPG_{p,q}(X; \rho_n)\).
\end{enumerate}

\begin{example}
\label{ex:intersect-curves}
Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$. 
This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
The community structure is also not clear from the adjacency matrix. 
However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).

@@

We performed two simulation experiments for algorithm \ref{alg:kcurves}.
In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}.
The setup is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(1/2, 1/2)\).
\item
  Draw \(t_1, ..., t_n \iid \Uniform(0, 1)\).
\item
  Let each \(x_i = g_{z_i}(t_i)\) where \(g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top\) and \(g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top\).
  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
\item
  Draw \(A \sim \RDPG(X; \rho_n)\) (for these simulations we fix \(\rho_n \equiv 1\)).
\end{enumerate}

For this simulation study, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community.
For each setting, we simulate 50 replicates.
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(0)} = p_2^{(0)} = (0, 0)\).
Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
The simulation results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance.

@@

\caption{Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.}\label{fig:sim-curves}
\end{figure}

@@

In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves.
The setup is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)\).
\item
  Draw \(t_1, ..., t_n \iid \Uniform(0, 1)\).
\item
  Let each \(x_i = g_{z_i}(t_i)\) where

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    \(g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top\)
  \item
    \(g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top\)
  \item
    \(g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top\).
  \end{enumerate}
\item
  Collect the latent vectors into matrix \(X = \big[ x_1 \mid \cdots \mid x_n \big]^\top\).
\item
  Draw \(A \sim \RDPG(X; \rho_n)\) (for these simulations we fix \(\rho_n \equiv 1\)).
\end{enumerate}

As in the previous setup, we set \(n = 128, 256, 512, 1024, 2048\), and we start with completely random initializations as well as initializations in which we know \(4\) or \(8\) labels from each community.
For each setting, we simulate 50 replicates.
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing \(p_1^{(k)} = (0, 0, 0)\) for each \(k\).
Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each \(n\) and \emph{a priori} known labels per community.
As in the previous \(K = 2\) setup, the results suggest that for sufficient \(n\), we do not benefit from having more known labels, but having a few results in greatly improved performance. Unlike in the previous setup, random initialization fails to improve with greater \(n\).

{\centering \includegraphics{draft_files/figure-latex/sim-curves-3-1}

\caption{Median and IQR error rates over 50 replicates in the simulation setup for $K = 3$ curves.}\label{fig:sim-curves-3}
\end{figure}

@@

The curve fitting step in algorithm \ref{alg:kcurves} requires estimation of some \(g: [0, 1] \to \mathbb{R}^d\) for each subset of points corresponding to each label.
One choice of curve is the Bezier polynomial.
The Bezier polynomial of order \(R\) is defined by function \(g\) with parameters \(p\) such that
\[g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}\]
where the Bezier coefficients \(p^{(r)}\) are vectors in \(\mathbb{R}^d\), and \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]\).

The general idea behind Bezier curve fitting to noisy data is as follows:
We wish to minimize the squared loss function
\begin{equation}
\label{eq:bezier-ls}
L(p, t; X) = \sum_i \|x_i - g(t_i; p)\|^2.
\end{equation}
Suppose that we know the timepoints \(t_1, ..., t_n\) which we wish to use to fit Bezier coefficients \(p^{(0)}, ..., p^{(R)}\) to vectors \(x_1, ..., x_n \in \mathbb{R}^d\) lying on or near a Bezier curve of order \(R\).
The squared loss then can be written as
\[L(p; t, X) = \|X - T p\|^2,\]
where \(p\) is defined as before, \(T\) is an \(n \times (R+1)\) matrix such that \(T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}\), and \(X = \big[ x_1 \mid \cdots \mid x_n \big]\).
Then the least squares estimates for the Bezier coefficients \(p\) is simply the linear least squares estimate
\begin{equation}
\label{eq:bezier-p-fit}
\hat{p} = (T^\top T)^{-1} T^\top X.
\end{equation}

On the other hand, suppose that we know the Bezier coefficients \(p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]^\top\), and we wish to fit timepoints \(t_1, ..., t_n\) to vectors \(x_1, ..., x_n\) lying on or near a Bezier polynomial of order \(R\).
In this setting, we can fit each individual \(t_i\) independently by minimizing \(L(t_i; p, x_i) = \|x_i - g(t_i; p)\|^2\) for each \(t_i\).
If \(g(t_i; p, x_i)\) is a polynomial of order \(R\), then \(\dot{L}(t_i; p, x_i)\) is also a polynomial, of order \(2 R - 1\).
Minimizing \(L(t_i; p, x_i)\) then simply involves finding the roots of \(\dot{L}(t_i; p, x_i)\) and choosing the root with the smallest objective value (in practice, the endpoints \(t_i = 0\) and \(t_i = 1\) are also checked).
For the Bezier curve of order \(R\), the full polynomial is given by:
\begin{equation}
\label{eq:dbezier}
\dot{L}(t_i; p, x_i) = \bigg(\sum_{r=0}^R \binom{R}{r} (-1)^r c_r t^r \bigg) \bigg(\sum_{r=0}^{R-1} \binom{R-1}{r} (-1)^r c_{r+1} t^r \bigg),
\end{equation}
where \(c_R = \sum_{r=0}^R (-1)^{R-r} \binom{R}{r} p^{(r)}\).
The full Bezier polynomial fitting method is detailed in algorithm \ref{alg:bezier-fit}.

\begin{algorithm}[h]
\label{alg:bezier-fit}
\DontPrintSemicolon
\SetAlgoLined
\setstretch{1.5}
\KwData{$x_1, ..., x_n \in \mathbb{R}^d$ lying on or near a Bezier curve of order $R$, stopping criterion $\epsilon$.}
\KwResult{Bezier coefficients $p$, timepoint values $t_1, ..., t_n$.}
Initialize timepoints $t_1, ..., t_n$ (e.g., by sampling from a random distribution).\;
\Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
  Fit $\hat{p}$ by equation \ref{eq:bezier-p-fit}.\;
  \For {$i = 1, ..., n$} {
    Fit $t_i$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimize equation \ref{eq:bezier-ls}.\;
  }
}
\caption{Procedure for fitting a Bezier polynomial to noisy data.}
\end{algorithm}

Algorithm \ref{alg:bezier-fit} is sensitive to initialization.
We propose two initialization strategies.
One involves using a one-dimensional Isomap \citep{Tenenbaum2000-ff} embedding normalized to the unit interval.
Since the vectors lie approximately on a one-dimensional manifold, a one-dimensional Isomap embedding should be able to approximate the manifold, and in particular the interpoint distances along the manifold, as long as the sample is dense along the manifold \citep{trosset2020rehabilitating}.

\section*{Appendix D: Supplementary Materials}
